---
title: "Kaplan–Meier estimator Notebook"
output: html_notebook
---

This notebook illustrates the effectiveness of survival analysis statistics to model event based processes.
The Kaplan-Meier estimator is used to obtain a survival function using a sample of random data generated from a weibull distribution. This script can be easily adapted for other distributions.

First the required R packages are loaded. 

```{r}
library(ggplot2)
library(dplyr)
library(survival)

```
In the below section are defined a number of functions. This first generates random data based on a (weibull) distribution function (data generation function). Next is a function which is used to display the obtained results and the theoretical expected values (results display function). The final 2 functions retrieve the period event rate and the cumulative event rate from the (weibull) distribution function.
```{r}
generate_deal <- function(nbr, end_period = 3650, horizon_period = 3650, minimum_duration = 1825, density_function = 0.0001) {
  # The deal period is a random period between the start period (= 1) and the end period
  deal_period <- sample(1:end_period, nbr, TRUE)
  # Initially we apply the same logic to the maturity period (so for the moment a deal can mature before it starts)
  maturity_period <- sample(1:end_period, nbr, TRUE)
  # Convert to a data frame
  data <- as.data.frame(cbind(end_period,deal_period,maturity_period))
  # The maturity period is increase with the deal period and the minimum duration
  data$maturity_period <- data$deal_period + data$maturity_period + minimum_duration
  # Without default the time value for the survival analysis is the minimum of the maturity period and the horizon period reduced with the deal period
  data$time <- data$maturity_period
  data$time[which(horizon_period < data$maturity_period)] <- horizon_period
  data$time <- data$time - data$deal_period
  # Without default all deals are censored for the survival analysis (set to 0 and NOT to 1)
  data$censor <- 0
  # Initially set all deals as not defaulted
  data$default_period <- -1
  ## Iterate over the data to determine a default period and time (for each time period until the deal time)
  for (time in 1:end_period) {
    # here we use a constant daily default rate (update with a function call to a weibull if time permits)
    default_rate <- density_function(period = time)
    default_check <- runif(n = nbr) # generates nbr uniform random variables from 0 to 1
    # if the default check is smaller than the default rate and the deal has not matured nor has it defaulted in an earlier period
    # then set the default period to the deal date increased with the current value of time variable we are iterating over
    data$default_period[which(default_check < default_rate & time < data$time & data$default_period == -1)] <- data$deal_period[which(default_check < default_rate & time < data$time & data$default_period == -1)] + time
    # update the censor status accordingly (defaulted = not censored)
    data$censor[which(data$default_period == data$deal_period + time)] <- 1 # set to 1 and NOT to 0
    # update the time value accordingly
    data$time[which(default_check < default_rate & time < data$time & data$default_period == data$deal_period + time)] <- time
  }
  return(data)
}

plot_km_vs_theoretical_value <- function(KM, horizon, cumulative_density_function) {
  plot(KM, xlim = c(0,horizon), ylim = c(0,1), col = 'blue', xaxs = 'r') # plots the KM with 95% confidence intervals
  par(new=TRUE)
  xi = seq(0:horizon)
  yi = cumulative_density_function(period = xi)
  plot(x = 0:horizon, y = 1 - yi, xlim = c(0,horizon), ylim = c(0,1), col = 'green',
       xlab = 'theoretical = green, Kaplan Meier = blue', xaxs = 'r', main = paste0('horizon set to ', horizon))
}

get_period_event_rate_weibull <- function(period, k = k_value, lambda = lambda_value) {
  p <- pweibull(period,k,scale = lambda)
  p_prev <- pweibull(period - 1,k,scale = lambda)
  return((p - p_prev) / (1 - p_prev))
}

get_period_cumulative_event_rate_weibull <- function(period, k = k_value, lambda = lambda_value) {
  return(pweibull(period,k,scale = lambda))
}
```
Below are listed the input parameters of the data generation process and of the distribution function. These values can be modified to examine the accuracy of the Kaplan-Meier estimator for other Weibull distributions. The corresponding Weibull distribution is displayed.
```{r}
end <- 100
horizon <- 100
k_value <- 2 # update this to get a different k value for the weibull rate
lambda_value <- 40 # update this to get a different lambda / scale value for the weibull rate
# generic functions to assign the specific event rate functions
period_event_rate <- get_period_event_rate_weibull
period_event_cumulative_rate <- get_period_cumulative_event_rate_weibull
# the weibull distribution plot
x <- seq(0, horizon, length=10*horizon)
y <- dweibull(x,k_value,scale = lambda_value) # density
data <- as.data.frame(cbind(x,y))
plot <- ggplot(data, aes(x = data$x, y = data$y))
print(plot + geom_line() + xlab(paste("k = ", k_value, " ; lambda = ", lambda_value)) + ylab("Weibull density"))
```
Below is the main section of the notebook.
First the data generation function is called (modify the nbr parameter to get a different sample size).
In the next 2 lines a Kaplan-Meier estimator model is generated.
In the last line the results display function is called which allows to get a feeling of the accuracy of the obtained survival function compared to the theoretical survival function.
```{r}
# generate data
data <- generate_deal(nbr = 40, end_period = end, horizon_period = horizon, minimum_duration = 10, density_function = period_event_rate)

# The below 2 lines of code are the only ones needed to construct the model
data_surv <- Surv(data$time,data$censor)
KM=survfit(data_surv~1) # Obtain the Kaplan-Meier estimator

# Kaplan Meier analysis
plot_km_vs_theoretical_value(KM = KM, horizon = end, cumulative_density_function = period_event_cumulative_rate)
```
